---
title: "Title: CSIT 165 Project 2"
output: html_document
author: "Names:  Daniel Chun, Hyun Lee"
date: "Date: `r Sys.Date()`"
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
---

### Introduction
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases.
*CSIT-165* disseminates data driven analyses to global decision makers.
*CSIT-165* is a conglomerate comprised of two fabricated entities: *World Health Organuzation (WHO)* and *US Pademic Response Team (USPRT)*. 

> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

```{r setup, include = FALSE}
# Loading packages:
library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
library(readr)
library(kableExtra) 
library(leaflet)
library(cowplot)
library(dplyr)
library(ggplot2)
library(tidyr)

# Loading data sets from RAW githubusercontent; curl and wget was not available
confirmed_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
death_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

confirmed_who_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"

death_who_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
# Download files to local; curl and wget not available
download.file(confirmed_url, basename(confirmed_url))
download.file(death_url, basename(death_url))
download.file(confirmed_who_url, basename(confirmed_who_url))
download.file(death_who_url, basename(death_who_url))


confirmed_us <- read_csv(url(confirmed_url))
death_us <- read_csv(url(death_url))
confirmed_who <-read_csv(url(confirmed_who_url))
death_who <-read_csv(url(death_who_url))

```

### Project Objectives

#### Objective 1:
> You are tasked to create a world map to gain an appreciation for where the most occurrences of COVID-19 confirmations and deaths are located.

>Create this map using leaflet for the most recent date as shown below. For this map, sum the confirmations and deaths of provinces into one value to depict the total number for the country they belong to. When creating a marker for each country in the map, calculate lat and long as the mean values for the provinces that make up each country.

>Customize the map to reflect the differences in magnitude for confirmations and deaths. In the example map below, circle markers that are blue represent low values, gray represents neutral values, and red represents high values. Low, middle, and high values were categorized to aesthetically map the markers based on their probabilistic distribution using the quartile function. You may use any method you like so that it is logical and allows visualization of value intensity. As well, customize the map to include hover labels that indicate country names and popup labels to show the value of confirmations and deaths for that country. For extra help using leaflet, consult this website along with the information provided in your textbooks.

```{r, include=TRUE, echo=TRUE}

```

#### Objective 2
> Seeing the global map of COVID-19 cases results in the stark realization that some countries are more affected than others. In order to narrow down your studies, create a table using kable from knitr listing the top countries as far as confirmations and deaths (sum values for provinces of the same country into one value and show the country only). Now that we are using RMarkdown to create HTML files, we have much more options for how we display our table. For reference on how to customize tables using knitr, visit this website. Consult the table below for an example of a customized table ranking cases by country. While it is not required to replicate this table exactly, it would be a fantastic challenge to show off your knitr prowess.

```{r, include=TRUE, echo=TRUE}
# sort death data
death_who %<>% select(-Lat, -Long)
death_who$sum_rows <- rowSums(death_who[,3:ncol(death_who)])
death_who %<>% select(1,2,ncol(death_who))
death_who %<>% group_by(`Country/Region`) %>% summarize(total_death = sum(sum_rows))
death_who %<>% arrange(desc(total_death))
death_who$Rank <- c(1:nrow(death_who))
death_who1 <- death_who[,c("Rank", "Country/Region", "total_death")]


# sort confirmation data
confirmed_who %<>% select(-Lat, -Long)
confirmed_who$sum_rows <- rowSums(confirmed_who[,3:ncol(confirmed_who)])
confirmed_who %<>% select(1,2,ncol(confirmed_who))
confirmed_who %<>% group_by(`Country/Region`) %>% summarize(total_confirmations = sum(sum_rows))
confirmed_who %<>% arrange(desc(total_confirmations))
confirmed_who$Rank <- c(1:nrow(confirmed_who))
confirmed_who1 <- confirmed_who[,c("Rank", "Country/Region", "total_confirmations")]

## Make tables
kbl(confirmed_who1, caption = "Confirmations") %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% scroll_box(width = "500px", height = "200px")

kbl(death_who1, caption = "Deaths") %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% scroll_box(width = "500px", height = "200px")




```

#### Objective 3
> After reading the top tables, you are stunned! The US overtakes every other country in terms of COVID-19 confirmations. As such, you are concerned about the state you live in and would like to understand how COVID-19 events have shaped the trajectory of the disease. Create two scatter plots to gain a better understanding. The first scatter plot should be California’s trajectory for confirmations. The second scatter plot should show California’s top three city trajectories for confirmations. You are interested in studying how the vaccine affected the number of confirmations. The Moderna vaccine was first available as an emergency use authorized (EUA) vaccine and required two shots spaced six weeks apart. Indicate on the plots the day the second dosage was given to those that received the first dosage the day Moderna was EUA (January 29th, 2021). As a diligent scientist that knows that new COVID variants have mutations in the spike protein (the region that the vaccine was developed for), you also want to study how confirmation rates change as new variants become the dominant infectious strain. Indicate on the plots when the delta and omicron variants became the dominant strain in California (May 11th, 2021 and November 26th, 2021 respectively). In the example below, the function plot_grid from the R package cowplot was to organize the graphs into a grid to more easily compare statewide vs top city plots.

```{r, include=TRUE, echo=TRUE}
## Isolate CA confirmations
confirmed_us %>% select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Country_Region, Lat, Long_, Combined_Key)) %>% filter(Province_State == "California") %>% data.frame() -> CA
CA %<>% select(-Province_State)
colnames(CA)<-gsub("X","",colnames(CA))
colnames(CA) <- as.Date(colnames(CA) , format = "%m.%d.%y")
CA_sum <- colSums(CA)
date <- colnames(CA)
ready <- data.frame(date,CA_sum)
ready$date <- as.Date(ready$date)

## Making graph for state
p <- ggplot(ready, aes(date,CA_sum)) + geom_point( color="blue") + theme_gray() + labs(x="Date", y="Confirmed cases", title = "COVID confirmations in CA")

## plotting vertical lines
p1 <- p + geom_vline(aes(xintercept=as.Date("2020-03-19")),color="red", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-01-29")),color="orange", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-07-04")),color="purple", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-12-20")),color="green", linetype = "longdash")

## labeling vertical lines
p2 <- p1 + geom_text(aes(as.Date("2020-03-19"), 7.0e+06, label ="HSA", hjust="right"))  + geom_text(aes(as.Date("2021-01-29"), 7.0e+06, label ="EUA + 6 WEEKS", hjust="right")) + geom_text(aes(as.Date("2021-07-04"), 7.0e+06, label ="Delta", hjust="right")) + geom_text(aes(as.Date("2021-12-20"), 7.0e+06, label ="Omicron", hjust="right"))

## adding tick marks
p3 <- p2 + scale_x_continuous(breaks = c(as.Date("2020-03-19"), as.Date("2021-01-29"), as.Date("2021-07-04"), as.Date("2021-12-20")), labels = c("2020-03-19", "2020-01-29", "2020-07-04", "2021-12-20"))  


confirmed_us %>% select(-c(UID, iso2, iso3, code3, FIPS, Country_Region, Lat, Long_, Combined_Key)) %>% filter(Province_State == "California") %>% data.frame() -> CA_city
CA_city$sum_rows <- rowSums(CA_city[,3:nrow(CA_city)])
CA_city %<>% arrange(desc(sum_rows)) %>% head(CA_city, n= 3)
CA_city %<>% select(-Province_State, -sum_rows)



## Plot for top 3 cities 
## Filter data frame for cities
CA_city %<>% pivot_longer(cols = starts_with("X"), names_to = "date", names_prefix = "X") %>% pivot_wider(names_from = Admin2, values_from = "value")
c <- mutate(CA_city, date = as.Date(date, format= "%m.%d.%y"))

## Plot for top 3 cities
top3 <- ggplot(c, aes(x=date)) + 
                                geom_line(aes(y=CA_city$`Santa Clara`, color="Santa Clara"), size=2) +
                                geom_line(aes(y=CA_city$`Los Angeles`, color="Los Angeles"), size=2) +
                                geom_line(aes(y=CA_city$`San Francisco`, color="San Francisco"), size=2)

c1 <- top3 + labs(x="Date", y="Confirmed cases", color="Cities") + theme(legend.position = "right")


## plotting vertical lines
c2 <- c1 + geom_vline(aes(xintercept=as.Date("2020-03-19")),color="red", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-01-29")),color="orange", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-07-04")),color="purple", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-12-20")),color="green", linetype = "longdash")

## labeling vertical lines
c3 <- c2 + geom_text(aes(as.Date("2020-03-19"), 2.0e+06, label ="HSA", hjust="right"))  + geom_text(aes(as.Date("2021-01-29"), 2.0e+06, label ="EUA + 6 WEEKS", hjust="right")) + geom_text(aes(as.Date("2021-07-04"), 2.0e+06, label ="Delta", hjust="right")) + geom_text(aes(as.Date("2021-12-20"), 2.0e+06, label ="Omicron", hjust="right"))

## adding tick marks
c4 <- c3 + scale_x_continuous(breaks = c(as.Date("2020-03-19"), as.Date("2021-01-29"), as.Date("2021-07-04"), as.Date("2021-12-20")), labels = c("2020-03-19", "2020-01-29", "2020-07-04", "2021-12-20"))  

## Use plot_grid
plot_grid(p3, c4, ncol = 1, align = "h")
                                

```

#### Objective 4
> Although these plots do not tell the whole story, they are great for helping us determine where to look. Different cities may have different populations, population densities, cultural discrepancies, compliance, and city regulations to name a few. We will explore the role of population on these metrics using visualizations. Arrange two scatter plots using cowplot’s plot_grid to show the relationship between population and confirmed counts as well as death counts and confirmed counts. You will need to use a log transform on all variables to show such a relationship. Please consult the example below for an idea of what this may look like. From these graphs we can see that population greatly affects confirmations and deaths. This coincides with our plots above as Los Angeles’s population is 301% greater than San Diego’s population and 406% greater than Riverside’s population!

```{r, include=TRUE, echo=TRUE}
## Population of CA cities
confirmed_us$sum_rows <- rowSums(confirmed_us[,12:ncol(confirmed_us)])
g1 <- data.frame(confirmed_us$Admin2, confirmed_us$sum_rows, death_us$Population)
gg <- ggplot(g1, aes(x=death_us.Population, y=confirmed_us.sum_rows)) + geom_point(color="red") + scale_x_continuous(trans = "log10") + scale_y_continuous(trans = "log10") + labs(x="US population", y="US Confirmed cases", title = "Graph 1: US Confirmed vs population")

## Make death vs confirmation for CA
death_us$sum_rows <- rowSums(death_us[,13:ncol(death_us)])
g2 <- data.frame(confirmed_us$sum_rows, death_us$sum_rows)
ggg <- ggplot(g2, aes(x=confirmed_us.sum_rows, y=death_us.sum_rows)) + geom_point(color="blue") + scale_x_continuous(trans = "log10") + scale_y_continuous(trans = "log10") + labs(x="US Confirmed cases", y="US Death", title = "Graph 2: US Confirmed vs Death")

## Use plot_grid
plot_grid(gg, ggg)

```

> Graph 1 graphs the population of each city across the country with the sum of confirmed cases from the past few months. Graph 2 has the sum of confirmed cases of each city as its x-axis and y-axis as the total sum of death per city. The purpose of this excercise is to demonstarte and further support the hypothesis that there is a correlation between the magnitude of population and confirmed cases/death. Graph 1 shows that has population increases, the confirmed cases increases. Graph 2 exhibits that has cases increase, so do the number of death. Therefore, it would be safe to conclude the hypothesis that the greater the population, the greater the death. Another indicator that universal healthcare is a necessity. 

#### GitHub repo: [csit165_project2](https://github.com/unleel/csit165_project2)
### GitHub Log
